Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests

### 2018-12-07 paper/sections/introduction.tex
more corrections

To this end, it was important to first undertake a thorough conceptual and
technical appraisal of our consumer data source.  In technical terms, screening
the information present in the `probe requests' and clustering them based on
their characteristics was essential in order to remove those emitted by devices
which do not indicate pedestrian activity, such as network enabled printers and
other fixed devices.  Related to this, a method to fingerprint Wi-Fi probes was
necessary in order to remove probes from individuals’ devices that in conceptual
terms should not be considered part of footfall; for instance when , an employee
is seated in an office within range of the sensor device.  A calibration of
sensor measurement was also essential on two grounds: first individuals may
carry multiple devices, or no device at all; and second, the positioning and
orientation of the sensor in the retail unit may lead to systematic over- or
under-enumeration.  These sources of bias in measurement must be accommodated by
manual recording of footfall at each location and the generalization of these
sample survey results to all locations and time periods.  As we describe in
detail below, manual validation of the data needed to be undertaken in parallel
with the technical profiling of the mix of consumer mobile devices that probed
our sensors, since the effectiveness of data cleaning procedures discussed in
this paper differ between individual locations and configurations.

>>>>

To this end, it was important to first undertake a thorough conceptual and
technical appraisal of our consumer data source. In technical terms, screening
the information present in the `probe requests' and clustering them based on
their characteristics was essential in order to remove those emitted by devices
which do not indicate pedestrian activity, such as network enabled printers and
other fixed devices. Related to this, a method to fingerprint Wi-Fi probes was
necessary in order to remove probes from individuals’ devices that in
conceptual terms should not be considered part of footfall; for instance when,
an employee is seated in an office within range of the sensor device. A
calibration of sensor measurement was also essential on two grounds: first
individuals may carry multiple devices, or no device at all; and second, the
positioning and orientation of the sensor in the retail unit may lead to
systematic over- or under-enumeration. These sources of bias in measurement
must be accommodated by manual recording of footfall at each location and the
generalization of these sample survey results to all locations and time
periods. As we describe in detail below, manual validation of the data needed
to be undertaken in parallel with the technical profiling of the mix of
consumer mobile devices that probed our sensors, since the effectiveness of
data cleaning procedures discussed in this paper differ between individual
locations and configurations.

### 2018-12-07 paper/sections/pilot_study.tex
more corrections

\begin{table}
\tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
{\begin{tabular}{lcc} 
	\toprule
	 Algorithm					& Time (s) 	& MAPE (\%) \\
	\midrule
	 Quantile					& 0.002 	&  27 \% \\
	 K-Means			 		& 0.007 	& -23 \% \\
	 Hierarchical Clustering	& 172.520 	&  -9 \% \\
	 Bagged Clustering 			& 0.135 	& -30 \% \\
	 Fisher 					& 3.034 	& -30 \% \\
	 Jenks Natural Break 		& 556.279 	& -30 \% \\
	 \bottomrule
\end{tabular}}
\label{classification-table}
\end{table}

To start we designed a small pilot study to validate the filtering and
clustering methodology against the scale and complexity of data collected on an
open public area such as a retail high street.  We also aim to find the
algorithm which is best suited for classification of signal strengths as 'low'
and 'high' to filter out the background noise.  The data was collected at Oxford
Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe
requests were collected using the sensor described in Chapter \ref{methodology}
and pedestrian footfall was manually recorded using the Android app
\citep{bala2018clicker}.  Being located at one of the busiest retail locations
in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe
requests during the half hour period, and 3,722 people were recorded manually
walking on the sidewalk during that time. The count was done at the location in
front of the store where the surveyor carried the sensor on a backpack for the
automated count and used a phone for doing the manual count. The sensor is kept
as close to the store window as possible and the manual count is done as a
cordon count in fron of the store.

As a first step we just aggregated the probe requests by their MAC address for
every minute to generate a minute by minute count of the number of people around
the sensor assuming each MAC address corresponds to a mobile device and hence a
pedestrian.  We then compare this preliminary `footfall' count to the actual
number of pedestrians recorded manually to check for the robustness. We use Mean
Absolute Percentage Error (MAPE) measure of robustness of the count since it
provides a simple and quick measurement while the street conditions ensure that
there are no intervals without any footfall.  We find that the MAPE in the raw
counts compared to the ground truth is around 425\%.  This suggests the presence
of large amount of noise in the data which might be generated due to sources of
uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need
for filtering the data.

We then classified the probe requests as ones with "high signal strength" and
"low signal strength" using various one dimensional clustering algorithms such
as k-means, quantile, hierarchical clustering, bagged clustering, fisher and
jenks natural breaks.  The results are shown in Table
\ref{classification-table}.  We found that while hierarchical clustering and
jenks gave us fairly low errors, they were too resource intensive for practical
use with a larger dataset.  We also found that k-means gave quickest results
with the lowest MAPE closely followed by quantile algorithm.  The cut-off point
or threshold for the collected data with which we could classify them as high
and low was -71 dBm.  We then removed all the probe requests which reported `low
signal strength' and repeated the same aggregation process as before to produce
footfall count.  This process resulted in a footfall count with a net MAPE of
30\%.  Though the results are encouraging we are still not completely confident
that our filtering process is indeed removing noise or has any correlation the
configuration of sensor or position of the mobile devices.  These need to be
addressed with a larger survey with multiple location of varying orientations.

@@

The next challenge was to identify probe requests which are generated by the
same device irrespective of MAC randomisation process.  We use the algorithm
defined in Chapter \ref{methodology} and assign a unique identifier or signature
to each probe request independent of the MAC address.  Since there is neither
prior research nor documentation on a universal behaviour of phones in
randomising their MAC addresses, we use the stationary device - the one used for
manual counting as a reference and find out the suitable time threshold,
$\alpha$ and threshold for sequence numbers, $\beta$ to be 16 seconds and 60
respectively via trial and error. This process is shown in Figure
\ref{pilot_clustering_params} This process is done on top the filtering done
based on signal strength and only for the probe requests with randomised MAC
addresses.  Figure \ref{pilot_results} shows the results of this clustering
process on a small set of randomised probe requests.  The probe request with
different randomised MAC address is shown by the colored points and the line
joining them shows the ones belonging to the same cluster hence expected to be
generated by the same device.  We finally aggregate the probe requests as before
but with the device signature rather than just MAC addresses this results in a
footfall count with a MAPE of -18\%.  A comparison of minute by minute counts
resulting from different filtering processes along with the ground truth is
shown in Figure \ref{pilot_results} showing the promising effectiveness of the
methods.

To conclude, from the pilot study we found that both filtering and clustering
methods we devices work on complex real world data and results in a final
pedestrian counts within a MAPE of 20\%. We also found `k-means' and `quantile'
are best algorithms for clustering signal strengths and the threshold for time
and sequence numbers for the clustering algorithm is around 16 and 60
respectively.

>>>>

begin{table}
\tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
{\begin{tabular}{lcc} 
	\toprule
	 Algorithm					& Time (s) 	& MAPE (\%) \\
	\midrule
	 Quantile					& 0.002 	&  27 \% \\
	 K-Means			 		& 0.007 	& -23 \% \\
	 Hierarchical Clustering	& 172.520 	&  -9 \% \\
	 Bagged Clustering 			& 0.135 	& -30 \% \\
	 Fisher 					& 3.034 	& -30 \% \\
	 Jenks Natural Break 		& 556.279 	& -30 \% \\
	 \bottomrule
\end{tabular}}
\label{classification-table}
\end{table}

To start, we designed a small pilot study to validate the filtering and
clustering methodology against the scale and complexity of data collected in an
open public area such as a retail high street. We also aimed to find the
algorithm which was best suited for the classification of signal strengths as
'low' and 'high' in order to filter out the background noise. The data was
collected at Oxford Street, London on 20 December 2017 from 12:30 to 13:00 hrs,
Wi-Fi probe requests were collected using the sensor described in Section
\ref{methodology} and pedestrian footfall was manually recorded using the
Android app - Clicker \citep{bala2018clicker}. Being located at one of the
busiest retail locations in the United Kingdom, the Wi-Fi sensor captured
approximately 60,000 probe requests during the half hour period; 3,722 people
were manually recorded walking on the pavement during that time. The surveyor
positioned himself at the front of a store while carrying the sensor in a
backpack and counted people walking by the store on the pavement (3m wide
approximately) using a mobile phone. The sensor was kept as close to the store
window as possible, and the manual count was done as a cordon count in front of
the store.

As a first step we aggregated the probe requests by their MAC addresses for
every minute to generate a minute by minute count of the number of people near
the sensor. We assumed that each MAC address corresponded to a mobile device and
hence a pedestrian. We then compared this preliminary `footfall' count to the
actual number of pedestrians recorded manually to check for it's robustness. We
used Mean Absolute Percentage Error (MAPE) as a measure of robustness of the
count, since it provided a simple and quick measurement and the street
conditions ensured that there are no intervals without any footfall. We found
that the MAPE in the raw counts compared to the ground truth was around 425\%.
This suggests the presence of large amount of noise in the data which may have
been generated by the sources of uncertainties discussed in Section
\ref{methodology} thus demonstrating the need for filtering the data.

We then classified the probe requests as `high signal strength' and `low signal
strength' using various one dimensional clustering algorithms such as k-means,
quantile, hierarchical clustering, bagged clustering, fisher and jenks natural
breaks. The results are shown in Table \ref{classification-table}. We found that
while hierarchical clustering and jenks gave us fairly low errors, they were too
resource intensive for practical use with a larger dataset. We also found that
k-means gave the quickest results with the lowest MAPE, closely followed by
quantile algorithm. The cut-off point or threshold for the collected data with
which we could classify as high and low was -71 dBm. We then removed all the
probe requests which reported `low signal strength' and repeated the same
aggregation process as before to produce footfall count. This process resulted
in a footfall count with a net MAPE of 30\%. Although the results are
encouraging we are still not completely confident that our filtering process is
removing noise or has any correlation the configuration of sensor or position of
the mobile devices. These concerns need to be addressed with a larger survey
with multiple locations of varying orientations.

@@

The next challenge was to identify the probe requests which are generated by the
same device irrespective of the MAC randomisation process. We used the algorithm
defined in Section \ref{methodology} and assigned a unique identifier or
signature to each probe request, independent of their the MAC addresses. Since
we didn't know the nature or frequency of the MAC address randomisation process,
we used the surveyor's mobile device as a reference. As the surveyor's device
was being actively used to count pedestrians and it's Wi-Fi module was kept
active without establishing connection to any network, the device was known to
be continuously probing for new networks. We also knew that the OUI of the
device was 'Google' and the device was regularly randomising it's MAC address,
thus providing us an excellent reference with which we could optimise the
parameters for our clustering algorithm. Using this reference device we observed
that the threshold for time $\alpha$ and the threshold for sequence numbers,
$\beta$ are 16 seconds and 60 respectively via trial and error. This process is
shown in Figure \ref{pilot_clustering_params}. This was undertaken on top the
filtering done based on signal strength, and only for the probe requests with
randomised MAC addresses. Figure \ref{pilot_results} shows the results of this
clustering process on a small set of randomised probe requests. The probe
requests with different randomised MAC address are shown by the coloured points
and the lines joining them show that those probe requests were most likely be
generated by the same device. We finally aggregated the probe requests as we did
before but with the device signature rather than MAC addresses. This results in
a footfall count with a MAPE of -18\% compared to the manual count. A comparison
of minute by minute counts resulting from different filtering processes along
with the ground truth is shown in Figure \ref{pilot_results} illustrating the
promising effectiveness of the methods.

To conclude, from the pilot study we found that both the filtering and the
clustering methods we devised worked on complex real world data and resulted in
final pedestrian counts within a MAPE of 20\%. We also found that `k-means' and
`quantile' are best algorithms for clustering signal strengths. Finally, we
observed that the best thresholds for time and sequence numbers in the
clustering algorithm is around 16 and 60 respectively.

### 2018-12-07 paper/sections/previous_work.tex
more corrections

In the past decade Wi-Fi has emerged as one of the most commonly used technology
in providing high speed internet access to mobile devices such as smartphones,
tablets and laptops in public and private spaces. This has resulted in multiple
Wi-Fi networks being available at almost every location in dense urban
environments. Traversing through this overlapping mesh of Wi-Fi networks, modern
mobile devices with Wi-Fi network interface regularly broadcast a special type
of signal known as `Probe Requests' in order to discover Wi-Fi networks
available to them.  This helps these devices to connect and switch between the
Wi-Fi networks seamlessly.

Probe requests are low level signals standardised by IEEE 802.11 specification
\citep{ieee2016} for service discovery and is implemented in any Wi-Fi capable
device irrespective of the manufacturer or the model. This ubiquity and
standardisation makes them an excellent source of open, passive, continuous,
and wireless data generated by Wi-Fi capable devices present at any given time
and location. Considering the unprecendented levels of mobile device ownership
in recent years, we can, in turn use this data to understand the population
distribution in highly dynamic urban environments with high spatial and
temporal granularity \citep{freud2015,konto2017}. While a Wi-Fi based method to
collect data offers us various advantages such as, easy scalability and
efficiency in terms of cost and time, It also introduces few systematic biases,
uncertainities in the collected data along with the serious risk of infringing
on the privacy of the mobile users. In this paper, using a set of probe
requests and manual counts collected at various high street locations across
London, we demonstrate that pedestrian footfall at these locations can be
estimated with considerable precision and accuracy while protecting the privacy
of the pedestrians.

Though, unlike GPS, the location of the Wi-Fi enabled mobile device cannot be
directly inferred from Wi-Fi, there are reliable methods to triangulate the
location of mobile devices from the locations of known access points(AP) and the
signal strength reported by them
\citep{he2003range,moore2004robust,lamarca2005place}. This can overcome the
usual shortcoming of GPS, which struggles for precision and accuracy in indoor
and densely built environments
\citep{zarim2006,kawaguchi2009wifi,xi2010locating}. Utilising this, we can
easily and quickly estimate trajectories of the mobile devices
\citep{Sorensenlocation,musa2012tracking} which can be used similar to the GPS
trajectories to understand individual travel patterns \citep{reki2007,Sap2015},
crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular
\citep{lu2010vehicle} and pedestrian movement
\citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}. Such data can
also be used in transportation planning and management to estimate travel time
\citep{musa2011wiflow} and real time traffic monitoring
\citep{abbott2013empirical}. Using techniques demonstrated by
\cite{franklin2006passive} and \cite{pang2007802} along with information present
in the probe requests one even model interactions between the users
\citep{cheng2012inferring,barbera2013signals,cunche2014know,cunche2014linking}
such as predicting which of them are most likely to meet again
\citep{cunche2012know}. Using the semantic information present in these probe
requests it even is possible to understand the nature of population at a large
scale \citep{di2016mind}.

Though extensive research has been carried out on this subject with feasible and
favorable results, in recent years, one of the major challenges faced in such
attempts has been the increasing attempt by mobile phone manufacturers to
protect their users’ privacy by anonymising the globally identifiable portion of
the probe requests, \citep{green2008}. Various methods have been devised to
overcome this anonymisation process such as estimating the device model
information from a known dataset of manufacturers and device behaviours
\citep{martin2016decomposition}; Scrambler attack using a small part of the
physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and
timing attack where the packet sequence information along with information
elements present in the probe request frame is used
\citep{matte2016,cheng2016can}. A combination of these methodologies has been
proven to produce de-anonymised globally unique device information
\citep{vanhoef2016,martin2017}. These approaches usually result in serious risk
of breach of privacy of the users of the mobile devices by revealing their
identifiable personal information.

There is a clear gap in research for exploring methodologies for estimating the
number of unique mobile devices from a set of anonymised probe requests, without
the need to reveal their original device information. Such technique has
various applications such as uncovering the urban wireless landscape
\citep{rose2010mapping}, revealing human activity at large scales
\citep{qin2013discovering}, estimating pedestrian numbers in crowds
\citep{schauer2014estimating,fukuzaki2015statistical} and even counting people
in hyper local scales such as queues \citep{wang2013measuring} With enough
infrastructure to collect such information we can even aim to generate a
real-time census of the city \citep{konto2017}. With this background we set out
to device and implement a methodology to reliably estimate human activity such
as pedestrian footfall from Wi-Fi probe requests without risking the breach of
privacy of the users involved.

>>>>

In the past decade, Wi-Fi has emerged as one of the most commonly used
technologies in providing high speed internet access to mobile devices such as
smartphones, tablets and laptops in public and private spaces. This has
resulted in multiple Wi-Fi networks being available at almost every location in
dense urban environments. Traversing through this overlapping mesh of Wi-Fi
networks, modern mobile devices with Wi-Fi network interfaces regularly
broadcast a special type of signal known as `Probe Requests' in order to
discover the Wi-Fi networks available to them. This helps these devices to
connect and switch between the Wi-Fi networks seamlessly.

Probe requests are low level signals standardised by IEEE 802.11 specification
\citep{ieee2016} for service discovery, and are implemented in any Wi-Fi
capable device irrespective of the manufacturer or the model. This ubiquity and
standardisation makes them an excellent source of open, passive, continuous,
and wireless data generated by Wi-Fi capable devices present at any given time
and location. Considering the unprecendented levels of mobile device ownership
in recent years, we can, in turn use this data to understand the population
distribution in highly dynamic urban environments with high spatial and
temporal granularity \citep{freud2015, konto2017}. While a Wi-Fi based method to
collect data offers us various advantages such as, easy scalability and
efficiency in terms of cost and time, it also introduces few systematic biases
and uncertainities in the collected data along with the serious risk of
infringing on the privacy of the mobile users. In this paper, using a set of
probe requests and manual counts collected at various high street locations
across London, we demonstrate that pedestrian footfall at these locations can
be estimated with considerable precision and accuracy while protecting the
privacy of the pedestrians.

Unlike GPS, the location of the Wi-Fi enabled mobile device cannot be directly
inferred from Wi-Fi, however there are reliable methods to triangulate the
location of mobile devices from the locations of known access points (AP) and
the signal strength reported by them \citep{he2003range, moore2004robust,
lamarca2005place}. This can overcome the usual shortcoming of GPS, which
struggles for precision and accuracy in indoor and densely built environments
\citep{zarim2006, kawaguchi2009wifi, xi2010locating}. Utilising this, we can
easily and quickly estimate trajectories of the mobile devices
\citep{Sorensenlocation, musa2012tracking} which can be used similary to the
GPS trajectories to understand individual travel patterns \citep{reki2007,
Sap2015}, crowd behaviour \citep{abedi2013bluetooth, mowafi2013tracking},
vehicular \citep{lu2010vehicle} and pedestrian movement
\citep{xu2013pedestrian, fukuzaki2014pedestrian, wang2016gait}. Such data can
also be used in transportation planning and management to estimate travel time
\citep{musa2011wiflow} and real time traffic monitoring
\citep{abbott2013empirical}. Using techniques demonstrated by
\cite{franklin2006passive} and \cite{pang2007802}, along with information
present in the probe requests, one can even model interactions between the
users \citep{cheng2012inferring, barbera2013signals, cunche2014know,
cunche2014linking} such as predicting which of them are most likely to meet
again \citep{cunche2012know}. Using the semantic information present in these
probe requests it even is possible to understand the nature of population at a
large scale \citep{di2016mind}.

Although extensive research has been carried out on this subject with feasible
and favorable results, in recent years, one of the major challenges faced in
such attempts has been the increasing attempt by mobile phone manufacturers to
protect their users’ privacy by anonymising the globally identifiable portion
of the probe requests \citep{green2008}. Various methods have been devised to
overcome this anonymisation process such as estimating the device model
information from a known dataset of manufacturers and device behaviours
\citep{martin2016decomposition}; Scrambler attack using a small part of the
physical layer specification for Wi-Fi \citep{vo2016, bloessl2015scrambler};
and timing attack where the packet sequence information along with information
elements present in the probe request frame is used \citep{matte2016,
cheng2016can}. A combination of these methodologies has been proven to produce
de-anonymised globally unique device information \citep{vanhoef2016,
martin2017}. These approaches usually result in serious risk of breach of
privacy of the users of the mobile devices by revealing their
identifiable personal information.

There is a clear gap in the research for exploring methodologies for estimating
the number of unique mobile devices from a set of anonymised probe requests,
without the need to reveal their original device information. Such a technique
has various applications such as uncovering the urban wireless landscape
\citep{rose2010mapping}, revealing human activity at large scales
\citep{qin2013discovering}, estimating pedestrian numbers in crowds
\citep{schauer2014estimating, fukuzaki2015statistical}, and even counting
people in hyper local scales such as queues \citep{wang2013measuring}. With
enough infrastructure to collect such information we can even aim to generate a
real-time census of the city \citep{konto2017}. With this background, we set
out to devise and implement a methodology to reliably estimate human activity
such as pedestrian footfall from Wi-Fi probe requests without risking a breach
of privacy of the users involved.
